---
title: "Class 1 - 7.10.2021"
author: "Alessandro Bramucci"
output:
  html_document:
    toc: true
    theme: united
---

# 1) R language essentials

```{r}
# R can be used as a calculator
2 + 2
exp(0)
log(1)

# Assigning a value to the object x
x <- 15

# To remove the object
rm(x)

# R handle data vectors as single objects. We can easily create data vectors
x <- c(1, 2, 3)

# We can check the names of the objects stored in the environment
ls()

# To remove all objects from the environment
rm(list = ls())

# R can easily create sequences of numbers
a <- 1:10 
b <- seq(from = 1, to = 10, by = 1)

# Generating random numbers from a normal distribution specifying mean and standard deviation; to obtain the same results we must set the same seed
set.seed(1234)

rnorm(n = 10, mean = 0, sd = 1)

# Assigning (a vector of) random numbers to an object called y
y <- rnorm(10, 0, 1)

# Length of the vector y
length(y)

# Logical operations. The operator "==" force R to make a "true or false" judgment
2 + 3 == 5
sqrt(25) == 4

# Other classical operators
100 >= 100
99 > 100
99 != 100
!(1==1) # One is not equal to one -> FALSE!
(1==1) | (2==3) # One is equal to one OR two is equal to three -> TRUE!
(1==1) & (2==3) # One is equal to one AND two is equal to three -> FALSE!
```

# 2) Matrix operations

R can perform standard matrix algebra operations. We can use matrix algebra functions in R to solve our problem from class.

$$
7x + 5y - 3z = 16 \\
3x -5y +2z	= -8 \\
5x +3y -7z = 0
$$

First, we rewrite the system using matrix and vector notation:

$$
\mathbf{A} = \left[\begin{array}
{rrr}
7 & 5 & -3 \\
3 & -5 & 2 \\
5 & 3 & -7 \\
\end{array}\right]
\mathbf{b} = \left[\begin{array}
{rrr}
x \\
y \\
z \\
\end{array}\right]
\mathbf{r} = \left[\begin{array}
{rrr}
16 \\
-8 \\
0 \\
\end{array}\right]
$$

In order to obtain the result vector *b*, we have to rearrange the model performing some simple matrix algebra operations.

$$
\mathbf{A}^{-1}\mathbf{A}\mathbf{b} = \mathbf{A}^{-1}\mathbf{r} \\
\text{remember that} \space \mathbf{A}^{-1}\mathbf{A} = \mathbf{I} \\
\mathbf{b} = \mathbf{A}^{-1}\mathbf{r}
$$ 

We are now ready to solve our system of equations using R:

```{r}
data <- c(7, 5, -3, 3, -5, 2, 5, 3, -7)

A <- matrix(data, nrow = 3, ncol = 3, byrow = TRUE)

r <- c(16, -8, 0)

b <- solve(A) %*% r

b
```
