---
title: "Estimation of Okun's coefficient for Germany, 1991-2018"
author: "Alessandro Bramucci"
date: "18.11.2021"
output:
  html_document:
    toc: true
    theme: united
---

# 1) Creating the variables for the regression

```{r, collapse = TRUE, message = FALSE, warning = FALSE}
# Load the data 
Okun <- read.csv("Okun.csv", 
                 skip = 1, 
                 sep = ";")

# Creating the first difference of the unemployment rate
Okun$dUnem <- c(NA, diff(Okun$Unem))

# Creating the first difference of real GDP
Okun$dGDP <- c(NA, diff(Okun$GDP))

# Creating lagged real GDP
library(dplyr)

Okun$lGDP <- lag(Okun$GDP, shift = 1)

# We now calculate the rate of growth of real GDP
Okun$GDP_gr <- Okun$dGDP/Okun$lGDP * 100

# Remove the first row
Okun <- Okun[-1,]

# Let's take a look at the data
head(Okun, 10L)

```

# 2) The linear regression

```{r, collapse = TRUE}
# Linear regression model. We regress unemployment on real GDP
reg1 <- lm(dUnem ~ GDP_gr, data = Okun)

# Scatter plot of the regression
plot(Okun$GDP_gr, Okun$dUnem,
     main = "Okun's Law with crisis in 2009",
     xlab = "Percent",
     ylab = "Percentage point")
abline(reg1, col = "blue", lty = 2, lwd = 2)
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)

# We remove 2009 (the year of the crisis) from the data
Okun2 <- subset(Okun, !(Okun$Year == 2009))

# We regress the change in unemployment rate on real GDP growth without the year of the crisis
reg2 <- lm(dUnem ~ GDP_gr, data = Okun2)

# We now want a scatter of our our new data (without the crisis)
plot(Okun2$GDP_gr, Okun2$dUnem,
     main = "Okun's Law without crisis in 2009",
     xlab = "Percent",
     ylab = "Percentage point")
abline(reg2, col = "blue", lty = 2, lwd = 2)
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)

# We can save the value of the estimated parameters for future computations
beta_0_2 <- unname(coef(reg2)["(Intercept)"])
beta_1_2 <- unname(coef(reg2)["GDP_gr"])
```

# 3) Regression tables

```{r, message = FALSE, collapse = TRUE}
library(stargazer)

# We now want to run a last model with a dummy that captures onyl the year of the crisis 2009
Okun$dummy_2009 <- 0

Okun$dummy_2009[Okun$Year == 2009] <- 1

reg3 <- lm(dUnem ~ GDP_gr + dummy_2009, data = Okun)
```

```{r}
stargazer(reg1, reg2, reg3, type = "text", keep.stat = c("n", "rsq"))
```

How much does real GDP has to grow per year in order to keep the change in the unemployment rate constant at 0%? First, we report the equation that we have just estimated.^[We will only use results from column (2). We have saved these values at the end of the previous section.]

$$dUnem = 0.624 - 0.421*GDP_{gr}$$
Since we want the change in unemployment from one year to the next to be equal to 0, we have to put 0 on the right-hand side of the estimated equation.

$$0 = 0.624 - 0.421*GDP_{gr}$$

Rearranging the terms, we get that GDP has to grow on average by `r round(beta_0_2/-beta_1_2, digits = 3)` to keep unemployment constant.

