---
title: "Standard deviation - example"
author: "Alessandro Bramucci"
date: "14.10.2021"
output: html_document
---

In this exercise, we simulate data for two variables drawing numbers from two normal distributions with same mean but different standard deviation. Goal of this simple exercise is to offer a visual understanding of the concept of standard deviation. We can think of these numbers as stock returns, for example.

Before we start, we must set the same "seed" using the *set.seed()* function if we want to obtain the same values from the simulations.

```{r}
set.seed(1234)
```

We draw numbers from two normal distributions with mean zero and standard deviation 10 and 5.

```{r}
stock_1 <- rnorm(100, mean = 0, sd = 10)
stock_2 <- rnorm(100, mean = 0, sd = 5)
```

We are now ready to plot the data.

```{r}

# Plot
plot(x = 1:100,
     y = stock_1,
     type = "l",
     col = "red",
     xlab = "Time",
     ylab = "Returns")
lines(x = 1:100,
      y = stock_2,
      type = "l",
      col = "blue")
legend(x = "topleft",
       legend = c("Stock 1", "Stock 2"),
       col = c("red", "blue"),
       bg = "transparent",
       bty = "n",
       lty = c(1,1),
       cex = 0.8)

```

To conclude, we can notice that the standard deviation of the variables *Stock 1* and *Stock 2* will be close, but not identical, to the standard deviation of the normal distribution from which data are drawn.

```{r}
sd(stock_1)
sd(stock_2)
```



